Simulation of Congenital Heart Defect Corrective Surgeries Using Thin Shell Elements

### 2012-03-02 Doc/Miccai/introduction.tex
spell check
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@388 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

In this paper we present an improved approach using a simulation model solely based on elastic thin shell elements. Our method is not limited to well-formed blood vessels and captures all low-level procedures necessary for aortoplasty. We use a relaxation process to arbitrarily join or attach blood vessels and patches without using springs that would add surplus energy to the physical system. By taking the overall tubular structure of blood vessels into consideration during generation of thin shell elements we achieve reasonable speed of the simulation.

>>>>

In this paper we present an improved approach for predictive simulation of congenital heart defect (CHD) corrective surgeries using a simulation model solely based on elastic thin shell elements. Our method is not limited to well-formed blood vessels and captures all low-level procedures necessary for aortoplasty. We use a relaxation process to arbitrarily join or attach blood vessels and patches without using springs that would add surplus energy to the physical system. By taking the overall tubular structure of blood vessels into consideration during generation of thin shell elements we achieve reasonable speed of the simulation.

### 2012-03-02 Doc/Miccai/results.tex
spell check
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@388 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

A quantitative validation by means of a comparision to real pre- and post-surgery image data is impossible at present. This is mainly due to ethical issues that arise during image data acquirement of infants, i.e. necessary heart rate lowering medications or exposure to radiation. There is no such data available that is obtained shortly after surgeries in infants where the growth of the patient would not interfere comparison. However, a qualitative comparison to real image data acquired months after a surgical intervention, which confirms a principal suitability of predictive simulations for surgery planning, can be found in \cite{Li2009}.

@@

\begin{figure}[tbh]
    \centering
    \begin{tabular}{cc}
     \includegraphics[width=0.3\columnwidth]{img/compare-bend.png}
      &
      \includegraphics[width=0.3\columnwidth]{img/compare-twist.png}
      \\
      \includegraphics[width=0.3\columnwidth]{img/compare-bend-other.png}
      &
      \includegraphics[width=0.3\columnwidth]{img/compare-twist-other.png}
    \end{tabular}
    \caption{Simulation of bending and twisting of an elestic tube. Top row: Thin shell element model; Bottom row: Real elastic tube manipulated by hand and Cosserat rod-based hybrid model \cite{Li2009} (Credits for the bottom images: Li et al.).}
    \label{fig-deformations}
\end{figure}

@@

We manually modeled an aortic arch based on real image data to be able to demonstrate suitability of our joining approach for different kinds of low-level surgical procedures on a concrete example. Therefore we chose a coarctation of an aortic arch that can be repaired by completely different surgical approaches that heavily alter the surface of the blood vessel \cite{Dodge2000}. Results can be examined in figure \ref{fig-surgery}. We show that our topological method for joining of shell elements through a controlled relaxation to their rest shape is able to handle different kinds of connections universally and in a correct manner.

@@

Finite element methods including thin shell elements have some constraints in common regarding the simulation mesh quality. A homogeneous neighborhood of equally sized elements is desirable. Especially when minimizing the number of elements, high mesh quality is necessary to maintain convergence and stability of the simulation. We showed that an elastic tubular structure can be simulated appropriately for CHD corrective surgeries with only a few elements. To allow for arbitrary incisions under such circumstances, the mesh must be remeshed in a way that its elements align along an incision. Mesh quality must be retained during remeshing and therfore the whole mesh may be affected. A limitation of our joining approach is that we need the same number of elements along connecting edges. Appropriate remeshing strategies must be developed to take full benefit of our proposed simulation approach. However, as a fallback, element count could be increased which will have negative impact on the simulation speed.

>>>>

A quantitative validation by means of a comparison to real pre- and post-surgery image data is impossible at present. This is mainly due to ethical issues that arise during image data acquirement of infants, i.e. necessary heart rate lowering medications or exposure to radiation. There is no such data available that is obtained shortly after surgeries in infants where the growth of the patient would not interfere comparison. However, a qualitative comparison to real image data acquired months after a surgical intervention, which confirms a principal suitability of predictive simulations for surgery planning, can be found in \cite{Li2009}.

@@

\begin{figure}[tbh]
    \centering
    \begin{tabular}{cc}
     \includegraphics[width=0.3\columnwidth]{img/compare-bend.png}
      &
      \includegraphics[width=0.3\columnwidth]{img/compare-twist.png}
      \\
      \includegraphics[width=0.3\columnwidth]{img/compare-bend-other.png}
      &
      \includegraphics[width=0.3\columnwidth]{img/compare-twist-other.png}
    \end{tabular}
    \caption{Simulation of bending and twisting of an elastic tube. Top row: Thin shell element model; Bottom row: Real elastic tube manipulated by hand and Cosserat rod-based hybrid model \cite{Li2009} (Credits for the bottom images: Li et al.).}
    \label{fig-deformations}
\end{figure}

@@

We manually modeled an aortic arch based on real image data to be able to demonstrate suitability of our joining approach for different kinds of low-level surgical procedures on a concrete example. Therefore we chose a coarctation of an aortic arch that can be repaired by several different surgical approaches that heavily alter the surface of the blood vessel \cite{Dodge2000}. Results can be examined in figure \ref{fig-surgery}. We show that our topological method for joining of shell elements through a controlled relaxation to their rest shape is able to handle different kinds of connections universally and in a correct manner.

@@

Finite element methods including thin shell elements have some constraints in common regarding the simulation mesh quality. A homogeneous neighborhood of equally sized elements is desirable. Especially when minimizing the number of elements, high mesh quality is necessary to maintain convergence and stability of the simulation. We showed that an elastic tubular structure can be simulated appropriately for CHD corrective surgeries with only a few elements. To allow for arbitrary incisions under such circumstances, the mesh must be remeshed in a way that its elements align along an incision. Mesh quality must be retained during remeshing and therefore the whole mesh may be affected. A limitation of our joining approach is that we need the same number of elements along connecting edges. Appropriate remeshing strategies must be developed to take full benefit of our proposed simulation approach. However, as a fall-back, element count could be increased which will have negative impact on the simulation speed.

### 2012-03-02 Doc/Miccai/shell-model.tex
spell check
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@388 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

In CHD corrective surgeries blood vessels respond to plastic interventions with a certain amount of deformation. Surgeons try to find an optimal approach to correct defects with a minimum of deformations and collisions that would have unnecessary negative impact on blood flow or that may lead to problems during growth of the patient. To be able to predict results for different surgical approaches a physicaly approriate model of blood vessels is mandatory. Furthermore, computation time of simulations must be low enough to enable interactive and usable planning software.

@@

Formulation of thin shell elements is based on the machanics of continuum. For a thorough description we refer the reader to a standard textbook about finite element methods (FEM), i.e. \cite{Reddy1993}. Thin shell elements are commonly used in physics and mechanics to model deformation of thin structures because in that case they are numerically more stable than volumetric FEMs. A shell element is a 2-dimensional element whose thickness is a parameter of the simulation. They combine forces from in-plane deformations, i.e. stretching or shearing, with bending forces (see figure \ref{fig-shell}). Such internal forces $\myvec{f}$ are a function of current displacement measured as difference of the current shape $\myvec{x}$ to the rest shape $\myvec{x_0}$. To accelerate computation, we use a corotational formulation that is based on the definition of local frames for each element.

In these local frames computation of internal forces is linear:
\begin{equation}
  \myvec{f} = \mathbf{K_e}(\myvec{x} - \myvec{x_0})
  \label{linear_static_model} 
\end{equation}
where $\mymat{K_e}$ is a stiffness matrix that is computed by Kirshohoff-Loves equations. This approach is similar to the model presented in  \cite{Comas2010b,Comas2010c}, although the interpolation functions we use are different to allow for better convergence\footnote{Our thin shell element model will be published separately}. We emphasize that the method described in this paper is independent of the concrete thin shell element model used and other non-corotational shell models can be used as well. When the simulation mesh is assembled, the contributions of each element are computed in a global frame using a different rotation matrix for each element:
\begin{equation}
\mathbb{K} = \mathcal{A}_{e=0}^N (\mathbf{R_e}^T \mathbf{K_e} \mathbf{R_e}) \hspace{10mm}
\mathbb{F} = \mathcal{A}_{e=0}^N ( \mathbf{R_e}^T \mathbf{K_e}(\myvec{x} - \myvec{x_0}))
\end{equation}
where $\mathcal{A}$ is the assembling operator over $N$ elements of the mesh, $\mathbb{K}$ is the global stiffness and $\mathbb{F}$ the internal forces of the entire shell element mesh structure. Using this approach, $\mathbb{K}$ is changing at each step of the simulation but the model is able to include geometric non-linear deformations. In practice, computation of node motion $\mathbb{U} = \mathbb{K}^{-1} \mathbb{F}$ is performed using a conjugate gradient algorithm.

>>>>

In CHD corrective surgeries blood vessels respond to plastic interventions with a certain amount of deformation. Surgeons try to find an optimal approach to correct defects with a minimum of deformations and collisions that would have unnecessary negative impact on blood flow or that may lead to problems during growth of the patient. To be able to predict results for different surgical approaches a physically appropriate model of blood vessels is mandatory. Furthermore, computation time of simulations must be low enough to enable interactive and usable planning software.

@@

Formulation of thin shell elements is based on the mechanics of continuum. For a thorough description we refer the reader to a standard textbook about finite element methods (FEM), i.e. \cite{Reddy1993}. Thin shell elements are commonly used in physics and mechanics to model deformation of thin structures because in that case they are numerically more stable than volumetric FEMs. A shell element is a 2-dimensional element whose thickness is a parameter of the simulation. They combine forces from in-plane deformations, i.e. stretching or shearing, with bending forces (see figure \ref{fig-shell}). Such internal forces $\myvec{f}$ are a function of current displacement measured as difference of the current shape $\myvec{x}$ to the rest shape $\myvec{x_0}$. To accelerate computation, we use a corotational formulation that is based on the definition of local frames for each element.

In these local frames computation of internal forces is linear:
\begin{equation}
  \myvec{f} = \mathbf{K_e}(\myvec{x} - \myvec{x_0})
  \label{linear_static_model} 
\end{equation}
where $\mymat{K_e}$ is a stiffness matrix that is computed by Kirchhoff-Love's equations. This approach is similar to the model presented in  \cite{Comas2010b,Comas2010c}, although the interpolation functions we use are different to allow for better convergence\footnote{Our thin shell element model will be published separately}. We emphasize that the method described in this paper is independent of the concrete thin shell element model used and other non-corotational shell models can be used as well. When the simulation mesh is assembled, the contributions of each element are computed in a global frame using a different rotation matrix for each element:
\begin{equation}
\mathbb{K} = \mathcal{A}_{e=0}^N (\mathbf{R_e}^T \mathbf{K_e} \mathbf{R_e}) \hspace{10mm}
\mathbb{F} = \mathcal{A}_{e=0}^N ( \mathbf{R_e}^T \mathbf{K_e}(\myvec{x} - \myvec{x_0}))
\end{equation}
where $\mathcal{A}$ is the assembling operator over $N$ elements of the mesh, $\mathbb{K}$ is the global stiffness and $\mathbb{F}$ the internal forces of the entire shell element mesh structure. Using this approach, $\mathbb{K}$ is changing at each step of the simulation but the model is able to include geometric non-linear deformations. In practice, computation of node motion $\mathbb{U} = \mathbb{K}^{-1} \mathbb{F}$ is performed using a conjugate gradient algorithm.
